---
title: "USL Championship Database — R Exploration"
subtitle: "Viewing tables, verifying edits, and querying with SQL and dplyr"
author: "USL Championship Data Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

---

# Quick Start: Connect and View Tables

This is the core workflow — run these three blocks any time you open RStudio
and want to inspect the database, especially after making changes with
`manage_db.py`.

**Step 1 — Install required packages** (only needed once per machine):

```{r install-packages, eval=FALSE}
install.packages(c("DBI", "RSQLite", "tidyverse", "dbplyr", "knitr", "kableExtra"))
```

**Step 2 — Load libraries:**

```{r load-libraries}
library(DBI)
library(RSQLite)
library(tidyverse)
library(dbplyr)
library(knitr)
library(kableExtra)
```

**Step 3 — Connect to the database:**

```{r connect}
# Update this path if your database file is in a different location.
# If this .Rmd file is in the same folder as the .db file, the line below works as-is.
db_path <- "usl_championship_26.db"

con <- dbConnect(RSQLite::SQLite(), dbname = db_path)
con
```

> Keep `con` open for the entire session. All queries below reuse this one
> connection. Close it with `dbDisconnect(con)` when you are done.

**Step 4 — Read any table into a data frame and view it:**

```{r read-any-table}
# Replace "players" with any table name: teams, season_rosters, player_snaps, etc.
players <- dbReadTable(con, "players")
players
```

```{r read-teams}
teams <- dbReadTable(con, "teams")
teams
```

```{r read-rosters}
season_rosters <- dbReadTable(con, "season_rosters")
season_rosters
```

> `df_print: paged` in this document's YAML means data frames render as
> interactive paged tables in RStudio. Click the column headers to sort,
> or use the page arrows to scroll through rows.

---

# Verifying Edits from manage_db.py

After running `manage_db.py` to add or delete a player or team, use the
patterns in this section to confirm the change took effect.

## Check a specific player by name

```{r verify-player-by-name}
player_name_search <- "Smith"   # change to the name you added or deleted

dbGetQuery(con, "
  SELECT player_id, name, position, dob, nation
  FROM players
  WHERE name LIKE ?
  ORDER BY name
", params = list(paste0("%", player_name_search, "%")))
```

## Check a specific team by name or abbreviation

```{r verify-team}
dbGetQuery(con, "
  SELECT team_id, name, abbreviation, city, conference, member_since
  FROM teams
  ORDER BY name
")
```

## Check the roster for a specific team and season

```{r verify-roster}
# Set these to the team_id and season you just edited.
check_team_id <- 111    # e.g. 111 = Louisville City FC
check_season  <- 2025

dbGetQuery(con, "
  SELECT p.player_id, p.name, p.position, sr.jersey_number, sr.is_captain
  FROM season_rosters sr
  JOIN players p ON sr.player_id = p.player_id
  WHERE sr.team_id = ? AND sr.season = ?
  ORDER BY p.name
", params = list(check_team_id, check_season))
```

## Row counts for all tables

A quick way to confirm something was added or removed:

```{r row-counts}
tables <- dbListTables(con)

tibble(table_name = tables) %>%
  rowwise() %>%
  mutate(rows = dbGetQuery(con, paste0("SELECT COUNT(*) AS n FROM [", table_name, "]"))$n) %>%
  ungroup() %>%
  arrange(desc(rows)) %>%
  kable(col.names = c("Table", "Row Count"), format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Reading All Tables into Data Frames

Use `dbReadTable()` to load a full table into R memory. This is appropriate
for the smaller reference and roster tables. For `player_snaps` (which grows
with every ingestion run), use a filtered query instead — see Section 4.

```{r read-all-tables}
players        <- dbReadTable(con, "players")
teams          <- dbReadTable(con, "teams")
stadiums       <- dbReadTable(con, "stadiums")
season_rosters <- dbReadTable(con, "season_rosters")
season_staff   <- dbReadTable(con, "season_staff")
season_stats   <- dbReadTable(con, "season_team_stats")
matches        <- dbReadTable(con, "matches")
appearances    <- dbReadTable(con, "appearances")
raw_snapshots  <- dbReadTable(con, "raw_snapshots")
```

Preview structure of any data frame with `glimpse()`:

```{r glimpse-example}
glimpse(players)
```

> **`dbReadTable()` vs. `dbGetQuery()`**
>
> - `dbReadTable(con, "tbl")` — loads the whole table, no filtering.
> - `dbGetQuery(con, "SELECT ...")` — runs any SQL and returns a data frame.
>   Use this whenever you want a subset, a join, or an aggregation.

---

# Discovering the Schema

## List all tables

```{r list-tables}
dbListTables(con)
```

## Column definitions for any table

`PRAGMA table_info()` returns column name, type, nullability, default, and
primary-key flag:

```{r schema-inspect}
get_schema <- function(con, tbl) {
  dbGetQuery(con, paste0("PRAGMA table_info('", tbl, "');")) %>%
    as_tibble() %>%
    select(cid, name, type, notnull, dflt_value, pk)
}

get_schema(con, "players") %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r schema-season-rosters}
get_schema(con, "season_rosters") %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Full CREATE TABLE statements

```{r full-ddl}
ddl <- dbGetQuery(con, "
  SELECT type, name, sql
  FROM sqlite_master
  WHERE sql IS NOT NULL
  ORDER BY
    CASE type WHEN 'table' THEN 1 WHEN 'view' THEN 2 ELSE 3 END,
    name
")

walk2(ddl$name, ddl$sql, ~ {
  cat(paste0("-- ", .x, "\n"))
  cat(.y, "\n\n")
})
```

---

# SQL Queries

## Simple SELECT

```{r sql-select}
dbGetQuery(con, "
  SELECT team_id, name, abbreviation, city, conference
  FROM teams
  ORDER BY conference, name
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Filtering with WHERE

```{r sql-where}
dbGetQuery(con, "
  SELECT player_id, name, nation, position
  FROM players
  WHERE position = 'Goalkeeper'
  ORDER BY name
") %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Aggregation with GROUP BY

```{r sql-group}
dbGetQuery(con, "
  SELECT nation, COUNT(*) AS player_count
  FROM players
  WHERE nation IS NOT NULL
  GROUP BY nation
  HAVING COUNT(*) >= 5
  ORDER BY player_count DESC
") %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## JOIN — teams and stadiums

```{r sql-join-stadium}
dbGetQuery(con, "
  SELECT t.name AS team, t.conference,
         s.name AS stadium, s.city AS stadium_city, s.capacity
  FROM teams t
  LEFT JOIN stadiums s ON t.stadium_id = s.stadium_id
  ORDER BY t.conference, t.name
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## JOIN — roster with player details

```{r sql-join-roster}
dbGetQuery(con, "
  SELECT p.name, p.position, sr.jersey_number, sr.is_captain,
         t.abbreviation AS team, sr.season
  FROM season_rosters sr
  JOIN players p ON sr.player_id = p.player_id
  JOIN teams   t ON sr.team_id   = t.team_id
  WHERE sr.season = 2025
  ORDER BY t.abbreviation, sr.jersey_number
  LIMIT 50
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Parameterized queries

Use `?` placeholders instead of pasting values into SQL strings:

```{r parameterized}
team_abbr <- "LOU"
season    <- 2025

dbGetQuery(con, "
  SELECT p.name, p.position, sr.jersey_number
  FROM season_rosters sr
  JOIN players p ON sr.player_id = p.player_id
  JOIN teams   t ON sr.team_id   = t.team_id
  WHERE t.abbreviation = ? AND sr.season = ?
  ORDER BY sr.jersey_number
", params = list(team_abbr, season)) %>%
  kable(caption = paste(season, "Roster:", team_abbr)) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Previewing a large table

For `player_snaps`, always filter before loading — it grows with every
ingestion run:

```{r preview-snaps}
dbGetQuery(con, "
  SELECT snap_date, team_id, player_id, games_played, goals, assists, mins_played
  FROM player_snaps
  ORDER BY snap_date DESC
  LIMIT 20
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# The Tidyverse Way: dbplyr

`dbplyr` translates `dplyr` verbs into SQL automatically. Nothing is pulled
into R until you call `collect()`.

## Create lazy table references

```{r dbplyr-refs}
tbl_teams      <- tbl(con, "teams")
tbl_players    <- tbl(con, "players")
tbl_rosters    <- tbl(con, "season_rosters")
tbl_snaps      <- tbl(con, "player_snaps")
tbl_matches    <- tbl(con, "matches")
tbl_appear     <- tbl(con, "appearances")
```

## See the SQL dbplyr generates

```{r dbplyr-show-query}
tbl_players %>%
  filter(position == "Goalkeeper") %>%
  select(player_id, name, nation) %>%
  arrange(name) %>%
  show_query()
```

## Aggregate with dplyr verbs

```{r dbplyr-nation-counts}
tbl_players %>%
  filter(!is.na(nation)) %>%
  count(nation, sort = TRUE) %>%
  filter(n >= 5) %>%
  collect() %>%
  kable(col.names = c("Nationality", "Players")) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Join tables

```{r dbplyr-join}
tbl_rosters %>%
  filter(season == 2025) %>%
  inner_join(tbl_players, by = "player_id") %>%
  inner_join(tbl_teams,   by = "team_id") %>%
  select(team = name.y, player = name.x, position, jersey_number) %>%
  arrange(team, jersey_number) %>%
  collect() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Aggregate on the database, then collect for plotting

```{r dbplyr-plot}
goals_by_team <- tbl_snaps %>%
  group_by(team_id) %>%
  summarise(total_goals = sum(goals, na.rm = TRUE), .groups = "drop") %>%
  inner_join(tbl_teams, by = "team_id") %>%
  select(team = name, total_goals) %>%
  arrange(desc(total_goals)) %>%
  collect()

ggplot(goals_by_team, aes(x = reorder(team, total_goals), y = total_goals)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Total Goals by Team (Player Snapshots)", x = NULL, y = "Goals") +
  theme_minimal()
```

> **When to `collect()`:** delay it as long as possible. Filter, join, and
> aggregate in the database first. Only pull into R when you need local
> operations (plotting, modeling, `stringr` functions, custom code).

---

# Data Quality Checks

## Foreign key integrity

```{r fk-inspect}
get_fks <- function(con, tbl_name) {
  dbGetQuery(con, paste0("PRAGMA foreign_key_list('", tbl_name, "');"))
}

fk_tables <- c("season_rosters", "appearances", "player_snaps", "matches", "teams")

map_dfr(fk_tables, ~ {
  fks <- get_fks(con, .x)
  if (nrow(fks) > 0) fks %>% mutate(source_table = .x, .before = 1)
}) %>%
  select(source_table, from_col = from, references_table = table, to_col = to) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Orphan record check

```{r orphan-check}
orphan_appearances <- dbGetQuery(con, "
  SELECT a.match_id, a.player_id
  FROM appearances a
  LEFT JOIN matches m ON a.match_id = m.match_id
  WHERE m.match_id IS NULL
")
cat("Orphan appearances (no matching match):", nrow(orphan_appearances), "\n")

orphan_snaps <- dbGetQuery(con, "
  SELECT ps.player_id, ps.team_id, ps.snap_date
  FROM player_snaps ps
  LEFT JOIN players p ON ps.player_id = p.player_id
  WHERE p.player_id IS NULL
")
cat("Orphan player_snaps (no matching player):", nrow(orphan_snaps), "\n")
```

---

# Cleanup

Always disconnect when you are finished:

```{r disconnect}
dbDisconnect(con)
cat("Connection closed.\n")
```

---

# Quick Reference

| Task | Code |
|---|---|
| Connect | `con <- dbConnect(RSQLite::SQLite(), "file.db")` |
| List tables | `dbListTables(con)` |
| Read whole table → data frame | `dbReadTable(con, "table_name")` |
| SQL query → data frame | `dbGetQuery(con, "SELECT ...")` |
| SQL with parameters | `dbGetQuery(con, "... WHERE x = ?", params = list(val))` |
| Column metadata | `dbGetQuery(con, "PRAGMA table_info('tbl')")` |
| Lazy dplyr reference | `tbl(con, "table_name")` |
| See generated SQL | `... %>% show_query()` |
| Pull lazy ref into R | `... %>% collect()` |
| Disconnect | `dbDisconnect(con)` |
