---
title: "Exploring the USL Championship Database in R"
subtitle: "A step-by-step tutorial for working with SQLite and relational data"
author: "USL Championship Data Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Introduction

This notebook walks through the **USL Championship SQLite database** from
the ground up. By running each code block in order you will learn how to:

1. Connect to a `.db` file from R using the `DBI` and `RSQLite` packages.
2. Discover what tables and views exist inside the database.
3. Read entire tables into data frames and inspect their structure.
4. Write SQL queries — from simple SELECTs to JOINs, aggregations, and window
   functions — and pull the results directly into R.
5. Use `dplyr` / `dbplyr` to query the database with tidyverse syntax *without*
   writing raw SQL.
6. Understand how foreign keys tie the tables together and how to navigate a
   relational schema.

> **Prerequisites:** Install the required packages before running any chunks.

```{r install-packages, eval=FALSE}
install.packages(c("DBI", "RSQLite", "tidyverse", "dbplyr", "knitr", "kableExtra"))
```

```{r load-libraries}
library(DBI)
library(RSQLite)
library(tidyverse)
library(dbplyr)
library(knitr)
library(kableExtra)
```

---

# 1 — Connecting to the Database

`DBI::dbConnect()` opens a connection to a SQLite file. If the file does not
exist, SQLite will silently create an empty database — so always double-check
the path.

```{r connect}
# Update this path to wherever your copy of the database lives.
db_path <- "usl_championship.db"

# Open the connection
con <- dbConnect(RSQLite::SQLite(), dbname = db_path)

# Quick sanity check — print the connection object
con
```

> **Tip:** Keep `con` open for the entire session. We will close it at the very
> end. Every query below reuses this single connection.

### Turning on foreign key enforcement

SQLite does not enforce foreign keys by default. If you plan to write data or
want constraint checks while exploring, flip the pragma on:

```{r fk-pragma}
dbExecute(con, "PRAGMA foreign_keys = ON;")
```

---

# 2 — Discovering the Schema

Before writing any queries, find out what is inside the database.

## 2.1 List all tables and views

```{r list-tables}
# Tables
tables <- dbListTables(con)
cat("Tables and views in the database:\n")
print(tables)
```

## 2.2 Row counts for every table

A quick way to gauge how much data each table holds:

```{r row-counts}
row_counts <- tibble(
  table_name = tables
) %>%
  rowwise() %>%
  mutate(
    rows = dbGetQuery(con, paste0("SELECT COUNT(*) AS n FROM [", table_name, "]"))$n
  ) %>%
  ungroup() %>%
  arrange(desc(rows))

row_counts %>%
  kable(col.names = c("Table / View", "Row Count"), format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 2.3 Inspect column definitions

`PRAGMA table_info()` returns the column name, type, nullability, default value,
and primary-key flag for any table.

```{r schema-inspect}
# Helper: return a tidy data frame of column info for one table
get_schema <- function(con, tbl) {
  dbGetQuery(con, paste0("PRAGMA table_info('", tbl, "');")) %>%
    as_tibble() %>%
    select(cid, name, type, notnull, dflt_value, pk)
}

# Example: look at the 'players' table
get_schema(con, "players") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

## 2.4 View the full schema DDL

For a deeper look, you can pull the original `CREATE TABLE` / `CREATE VIEW`
statements:

```{r full-ddl}
ddl <- dbGetQuery(con, "
  SELECT type, name, sql
  FROM sqlite_master
  WHERE sql IS NOT NULL
  ORDER BY
    CASE type WHEN 'table' THEN 1 WHEN 'view' THEN 2 ELSE 3 END,
    name
")

# Print each statement
walk2(ddl$name, ddl$sql, ~ {
  cat(paste0("-- ", .x, "\n"))
  cat(.y, "\n\n")
})
```

---

# 3 — Reading Tables into R

## 3.1 The quick way: `dbReadTable()`

`dbReadTable()` pulls an entire table into a data frame. This is fine for
small reference tables but be cautious with large tables — you are loading
everything into memory at once.

```{r read-teams}
teams <- dbReadTable(con, "teams")
glimpse(teams)
```

```{r read-stadiums}
stadiums <- dbReadTable(con, "stadiums")
glimpse(stadiums)
```

```{r read-players}
players <- dbReadTable(con, "players")
glimpse(players)
```

## 3.2 Preview large tables first

For bigger tables like `player_snaps` and `matches`, preview before committing
to a full read:

```{r preview-snaps}
# Fetch just the first 10 rows
dbGetQuery(con, "SELECT * FROM player_snaps LIMIT 10") %>%
  glimpse()
```

```{r read-matches}
matches <- dbReadTable(con, "matches")
glimpse(matches)
```

```{r read-appearances}
appearances <- dbReadTable(con, "appearances")
glimpse(appearances)
```

> **When to use `dbReadTable()` vs. `dbGetQuery()`:**
>
> - `dbReadTable(tbl)` — grab the whole table (convenient, but no filtering).
> - `dbGetQuery(sql)` — run any SQL and return a data frame (full control).
>
> A good rule of thumb: if the table has fewer than ~50 000 rows and you want
> all of it, `dbReadTable()` is fine. Otherwise, filter server-side with SQL.

---

# 4 — Querying with SQL

This section walks through increasingly complex SQL patterns, all executed
against the live database.

## 4.1 Simple SELECT

Retrieve all 24 teams sorted by conference and city:

```{r sql-select}
dbGetQuery(con, "
  SELECT team_id, name, city, abbreviation, conference
  FROM teams
  ORDER BY conference, city
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 4.2 Filtering with WHERE

Find all goalkeepers in the database:

```{r sql-where}
dbGetQuery(con, "
  SELECT player_id, name, nation, position
  FROM players
  WHERE position = 'Goalkeeper'
  ORDER BY name
  LIMIT 15
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

## 4.3 Aggregation: GROUP BY and HAVING

Count the number of players by nationality, showing only countries with 5+
players:

```{r sql-group}
dbGetQuery(con, "
  SELECT nation, COUNT(*) AS player_count
  FROM players
  WHERE nation IS NOT NULL
  GROUP BY nation
  HAVING COUNT(*) >= 5
  ORDER BY player_count DESC
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

## 4.4 JOINs — connecting related tables

### Teams and their stadiums

```{r sql-join-stadium}
dbGetQuery(con, "
  SELECT t.name AS team, t.conference,
         s.name AS stadium, s.city AS stadium_city
  FROM teams t
  LEFT JOIN stadiums s ON t.stadium_id = s.stadium_id
  ORDER BY t.conference, t.name
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Top scorers with team names

This joins the `v_player_summaries` view to `players` and `teams` to show
human-readable names alongside the stats:

```{r sql-top-scorers}
dbGetQuery(con, "
  SELECT p.name AS player,
         t.abbreviation AS team,
         ps.goals,
         ps.assists,
         ps.games_played,
         ROUND(CAST(ps.goals AS REAL) / NULLIF(ps.games_played, 0), 2)
           AS goals_per_game
  FROM v_player_summaries ps
  JOIN players p ON ps.player_id = p.player_id
  JOIN teams t   ON ps.team_id   = t.team_id
  WHERE ps.goals > 0
  ORDER BY ps.goals DESC, ps.assists DESC
  LIMIT 20
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 4.5 Using views

The database ships with three prebuilt views. They work exactly like tables
in a `SELECT` statement:

```{r sql-standings}
standings <- dbGetQuery(con, "SELECT * FROM v_standings")
standings %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r sql-rosters-sample}
# Current roster for one team
dbGetQuery(con, "
  SELECT team_name, player_name, position, jersey_number, is_captain
  FROM v_current_rosters
  WHERE abbreviation = 'LOU'
  ORDER BY jersey_number
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

> **What is a view?** A view is a saved query — it doesn't store any data.
> Every time you `SELECT` from it, the underlying query runs against the real
> tables. This means views are always up-to-date but can be slower than a
> materialized table for very heavy queries.

## 4.6 Subqueries and CASE expressions

Find each team's home and away records:

```{r sql-home-away}
dbGetQuery(con, "
  SELECT
    t.name,
    SUM(CASE WHEN m.home_team_id = t.team_id AND m.home_score > m.away_score
             THEN 1 ELSE 0 END) AS home_wins,
    SUM(CASE WHEN m.away_team_id = t.team_id AND m.away_score > m.home_score
             THEN 1 ELSE 0 END) AS away_wins,
    COUNT(*) AS total_matches
  FROM teams t
  JOIN matches m ON t.team_id IN (m.home_team_id, m.away_team_id)
  WHERE m.competition = 'League'
  GROUP BY t.team_id
  HAVING total_matches > 0
  ORDER BY home_wins + away_wins DESC
") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 5 — The Tidyverse Way: dbplyr

`dbplyr` lets you write `dplyr` verbs (`filter`, `select`, `mutate`,
`summarise`, `left_join`, etc.) that are automatically translated into SQL
and run on the database. Nothing is pulled into R until you call `collect()`.

## 5.1 Creating lazy table references

```{r dbplyr-refs}
# tbl() creates a lazy reference — no data is fetched yet
tbl_teams      <- tbl(con, "teams")
tbl_players    <- tbl(con, "players")
tbl_matches    <- tbl(con, "matches")
tbl_snaps      <- tbl(con, "player_snaps")
tbl_rosters    <- tbl(con, "season_rosters")
tbl_appear     <- tbl(con, "appearances")
```

## 5.2 Inspect the generated SQL

One of the best features of `dbplyr`: you can see the SQL it writes for you
using `show_query()`.

```{r dbplyr-show-query}
tbl_players %>%
  filter(position == "Forward") %>%
  select(player_id, name, nation) %>%
  arrange(name) %>%
  head(10) %>%
  show_query()
```

## 5.3 Aggregation with dplyr verbs

```{r dbplyr-nation-counts}
tbl_players %>%
  filter(!is.na(nation)) %>%
  count(nation, sort = TRUE) %>%
  filter(n >= 5) %>%
  collect() %>%
  kable(col.names = c("Nationality", "Players")) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

## 5.4 Joining tables the tidy way

```{r dbplyr-join}
tbl_rosters %>%
  filter(season == 2025, is_captain == 1) %>%
  inner_join(tbl_players, by = "player_id") %>%
  inner_join(tbl_teams, by = "team_id") %>%
  select(team = name.y, captain = name.x, position, jersey_number) %>%
  arrange(team) %>%
  collect() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 5.5 When to `collect()`

`collect()` executes the query and pulls the result into a local tibble. Key
guidelines:

- **Delay** `collect()` as long as possible — let the database do the filtering,
  joining, and aggregating.
- **Call** `collect()` when you need the data in R for plotting, modeling, or
  any operation that `dbplyr` can't translate to SQL (e.g., string
  manipulations with `stringr`, custom functions).

```{r dbplyr-collect-example}
# Build a pipeline entirely on the DB, then collect for plotting
goals_by_team <- tbl_snaps %>%
  group_by(team_id) %>%
  summarise(total_goals = sum(goals, na.rm = TRUE), .groups = "drop") %>%
  inner_join(tbl_teams, by = "team_id") %>%
  select(team = name, total_goals) %>%
  arrange(desc(total_goals)) %>%
  collect()

# Now plot in R
ggplot(goals_by_team, aes(x = reorder(team, total_goals), y = total_goals)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Total Goals by Team (Player Snapshots)",
    x = NULL, y = "Goals"
  ) +
  theme_minimal()
```

---

# 6 — Understanding the Relational Model

## 6.1 Entity-relationship overview

The database is organized in four tiers:
 
| Tier | Tables | Update Frequency |
|------|--------|------------------|
| **Reference** | `stadiums`, `teams`, `players` | Rarely changes |
| **Season** | `season_rosters`, `season_staff`, `season_team_stats` | Annually |
| **Match** | `matches`, `appearances` | Weekly during season |
| **Snapshot** | `raw_snapshots`, `player_snaps` | Per ingestion run |

The key relationships are:

- `teams.stadium_id` → `stadiums.stadium_id` (each team has a home stadium)
- `season_rosters.team_id` → `teams.team_id` (roster assignment)
- `season_rosters.player_id` → `players.player_id` (which player)
- `matches.home_team_id` / `away_team_id` → `teams.team_id`
- `appearances.match_id` → `matches.match_id`
- `appearances.player_id` → `players.player_id`
- `player_snaps.team_id` → `teams.team_id`
- `player_snaps.player_id` → `players.player_id`

## 6.2 Exploring foreign keys

SQLite stores foreign key metadata that you can query directly:

```{r fk-inspect}
# List foreign keys for a given table
get_fks <- function(con, tbl_name) {
  dbGetQuery(con, paste0("PRAGMA foreign_key_list('", tbl_name, "');"))
}

# Show FKs for several tables
fk_tables <- c("teams", "matches", "appearances", "player_snaps", "season_rosters")

map_dfr(fk_tables, ~ {
  fks <- get_fks(con, .x)
  if (nrow(fks) > 0) {
    fks %>% mutate(source_table = .x, .before = 1)
  }
}) %>%
  select(source_table, from_col = from, references_table = table, to_col = to) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 6.3 Checking for orphan records

A common data-quality check: do any child records reference parent IDs that
don't exist?

```{r orphan-check}
# Are there any appearances referencing a match_id not in matches?
orphan_appearances <- dbGetQuery(con, "
  SELECT a.match_id, a.player_id
  FROM appearances a
  LEFT JOIN matches m ON a.match_id = m.match_id
  WHERE m.match_id IS NULL
")
cat("Orphan appearances (no matching match):", nrow(orphan_appearances), "\n")

# Any player_snaps referencing a player_id not in players?
orphan_snaps <- dbGetQuery(con, "
  SELECT ps.player_id, ps.team_id, ps.snap_date
  FROM player_snaps ps
  LEFT JOIN players p ON ps.player_id = p.player_id
  WHERE p.player_id IS NULL
")
cat("Orphan player_snaps (no matching player):", nrow(orphan_snaps), "\n")
```

---

# 7 — Parameterized Queries

Hard-coding values into SQL strings invites SQL injection and makes code
brittle. Use parameterized queries instead:

```{r parameterized}
# Safe: the ? placeholder is filled in by dbGetQuery's params argument
team_abbr <- "LOU"

roster <- dbGetQuery(con, "
  SELECT p.name, p.position, sr.jersey_number
  FROM season_rosters sr
  JOIN players p ON sr.player_id = p.player_id
  JOIN teams t   ON sr.team_id   = t.team_id
  WHERE t.abbreviation = ? AND sr.season = ?
  ORDER BY sr.jersey_number
", params = list(team_abbr, 2025))

roster %>%
  kable(caption = paste("2025 Roster:", team_abbr)) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

> **Why parameterize?** Even in an analysis notebook you should build the habit.
> The database driver handles type conversion, quoting, and escaping for you,
> which eliminates a whole class of bugs.

---

# 8 — Practical Analysis Examples

## 8.1 Player snapshot deep dive

Pull the full stat line for a single player across all their snapshots to see
how cumulative stats grow over the season:

```{r player-timeline}
# Pick a player with multiple snapshots
multi_snap_player <- dbGetQuery(con, "
  SELECT player_id, COUNT(*) as snap_count
  FROM player_snaps
  GROUP BY player_id
  HAVING snap_count > 1
  ORDER BY snap_count DESC
  LIMIT 1
")

if (nrow(multi_snap_player) > 0) {
  pid <- multi_snap_player$player_id[1]

  timeline <- dbGetQuery(con, "
    SELECT ps.snap_date, p.name, ps.games_played, ps.goals, ps.assists,
           ps.mins_played, ps.shots_on_target
    FROM player_snaps ps
    JOIN players p ON ps.player_id = p.player_id
    WHERE ps.player_id = ?
    ORDER BY ps.snap_date
  ", params = list(pid))

  timeline %>%
    kable(caption = paste("Snapshot Timeline:", timeline$name[1])) %>%
    kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
} else {
  cat("No players with multiple snapshots found.\n")
}
```

## 8.2 Match results heatmap data

Prepare a home-vs-away matrix of results:

```{r match-matrix}
result_matrix <- dbGetQuery(con, "
  SELECT
    ht.abbreviation AS home,
    at.abbreviation AS away,
    m.home_score,
    m.away_score,
    m.date
  FROM matches m
  JOIN teams ht ON m.home_team_id = ht.team_id
  JOIN teams at ON m.away_team_id = at.team_id
  WHERE m.competition = 'League'
  ORDER BY m.date
")

result_matrix %>%
  mutate(result_label = paste0(home_score, "-", away_score)) %>%
  select(date, home, away, result_label) %>%
  kable(caption = "League Match Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 8.3 Position distribution across the league

```{r position-dist}
pos_data <- dbGetQuery(con, "
  SELECT p.position, COUNT(*) AS n
  FROM players p
  WHERE p.position IS NOT NULL
  GROUP BY p.position
  ORDER BY n DESC
")

ggplot(pos_data, aes(x = reorder(position, n), y = n)) +
  geom_col(fill = "coral") +
  coord_flip() +
  labs(title = "Players by Position", x = NULL, y = "Count") +
  theme_minimal()
```

---

# 9 — Tips for Working with Relational Databases in R

1. **Always close your connection** when you are done (`dbDisconnect(con)`).
   Open connections hold file locks and consume resources.

2. **Let the database do the heavy lifting.** Filter, join, and aggregate in
   SQL (or `dbplyr`) before pulling data into R. Transferring millions of rows
   only to `filter()` them locally is wasteful.

3. **Use `dbplyr` for exploration, raw SQL for production.** `dbplyr` is
   excellent for iterative analysis because you can build queries incrementally.
   For final reports or scheduled jobs, explicit SQL gives you full control and
   is easier to optimize.

4. **Watch out for type coercion.** SQLite is loosely typed — a column declared
   as `INTEGER` can still contain text. R's `dbReadTable()` will guess types,
   but you may need to cast manually with `as.integer()`, `as.Date()`, etc.

5. **Beware of `NULL` vs. `NA`.** SQL `NULL` becomes R `NA` when fetched, but
   the behavior inside SQL (`IS NULL`, `COALESCE`, `NULLIF`) is different from
   R's `is.na()` and `replace_na()`. Be explicit on both sides.

6. **Use transactions for writes.** If you are inserting or updating data,
   wrap your operations in `dbBegin()` / `dbCommit()` (or `dbRollback()` on
   failure) to keep the database consistent.

7. **Index columns you filter on.** If you repeatedly query by `team_id` or
   `player_id`, adding an index speeds things up enormously for larger datasets:
   `dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_snaps_player ON player_snaps(player_id)")`.

8. **Composite primary keys matter.** Several tables here — `season_rosters`,
   `appearances`, `player_snaps` — use multi-column primary keys. When joining
   or deduplicating, you need *all* key columns, not just one.

---

# 10 — Cleanup

Always disconnect when you are finished:

```{r disconnect}
dbDisconnect(con)
cat("Connection closed.\n")
```

---

# Appendix: Quick Reference

| Task | Function |
|------|----------|
| Connect | `dbConnect(RSQLite::SQLite(), "file.db")` |
| List tables | `dbListTables(con)` |
| Read whole table | `dbReadTable(con, "table_name")` |
| Run SQL → data frame | `dbGetQuery(con, "SELECT ...")` |
| Run SQL (no result) | `dbExecute(con, "INSERT ...")` |
| Lazy dplyr reference | `tbl(con, "table_name")` |
| See generated SQL | `... %>% show_query()` |
| Pull lazy → local | `... %>% collect()` |
| Column metadata | `dbGetQuery(con, "PRAGMA table_info('tbl')")` |
| Foreign keys | `dbGetQuery(con, "PRAGMA foreign_key_list('tbl')")` |
| Disconnect | `dbDisconnect(con)` |
