---
title: "USL Championship — CSV Ingestion Pipeline"
subtitle: "Convert, review, and import snapshot data into the SQLite database"
author: "USL Championship Data Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook ingests data from `.csv` files (converted from `.xlsx` exports)
into the `usl_championship.db` SQLite database. **Run each section in order,
one chunk at a time in RStudio.** Do not knit the whole document in one shot —
Section 3 requires you to review a table and edit a vector before continuing.

The pipeline covers:

1. **Convert** — turn any new `.xlsx` files into `.csv`
2. **Ingest named CSVs** — sync `matches`, `players`, `appearances`, etc.
3. **Review new player names** — inspect unmatched names from the numbered
   snapshot CSVs before adding them to the `players` table
4. **Add approved players** — insert only the names you have validated
5. **Ingest snapshot CSVs** — load per-team, per-date stat files into
   `raw_snapshots` and `player_snaps`
6. **Audit** — verify counts and see what (if anything) was still skipped

---

# 0 — Setup

```{r install-packages, eval=FALSE}
install.packages(c(
  "DBI", "RSQLite", "readxl", "readr", "dplyr",
  "stringi", "lubridate", "knitr", "kableExtra"
))
```

```{r load-libraries}
library(DBI)
library(RSQLite)
library(readxl)
library(readr)
library(dplyr)
library(stringi)
library(lubridate)
library(knitr)
library(kableExtra)
```

## Paths and connection

**Edit `base` and `db_path` to match wherever your files live.**

```{r config}
base    <- "/home/jaycoughlon/Desktop/usl_analytics/"
db_path <- paste0(base, "usl_championship.db")

con <- dbConnect(RSQLite::SQLite(), dbname = db_path)
cat("Connected to:", db_path, "\n")
```

---

## Helper functions

These are used throughout the notebook. Run this chunk once and leave it.

```{r helpers}
# ── Logging ──────────────────────────────────────────────────────────────────
log_msg <- function(...) cat(format(Sys.time(), "[%H:%M:%S]"), ..., "\n")

# ── Safe insert: append only if there are rows to add ────────────────────────
safe_insert <- function(con, table, df) {
  if (nrow(df) == 0) return(invisible(0L))
  dbWriteTable(con, table, df, append = TRUE)
  invisible(nrow(df))
}

# ── Convert "First Last" → "Last, First" (handles II/III/Jr./Sr. suffixes) ──
to_last_first <- function(name) {
  suffixes <- c("II", "III", "IV", "Jr.", "Sr.", "Jr", "Sr")
  parts    <- strsplit(trimws(name), " ")[[1]]
  suffix   <- ""
  if (length(parts) > 1 && parts[length(parts)] %in% suffixes) {
    suffix <- paste0(" ", parts[length(parts)])
    parts  <- parts[-length(parts)]
  }
  if (length(parts) == 1) return(paste0(parts[1], suffix))
  last  <- parts[length(parts)]
  first <- paste(parts[-length(parts)], collapse = " ")
  paste0(last, suffix, ", ", first)
}

# ── Build a name-lookup table from the DB (all players) ──────────────────────
# Returns a data frame with columns: player_id, name, name_fl, name_norm
# name_fl   = "First Last" version of the DB's "Last, First" format
# name_norm = accent-stripped name_fl for fuzzy matching
build_name_lookup <- function(con) {
  db <- dbGetQuery(con, "SELECT player_id, name FROM players")
  db$name_fl   <- sapply(strsplit(db$name, ", "),
                         function(x) if (length(x) == 2) paste(x[2], x[1]) else x[1])
  db$name_norm <- stri_trans_general(db$name_fl, "Latin-ASCII")
  db
}

# ── Name overrides: CSV spelling → known player_id ───────────────────────────
# Add new entries here whenever a CSV uses a nickname, shortened name, or typo
# for a player who already exists in the DB.
manual_overrides <- c(
  "Samuel McIllhatton"    = 6005L,  # DB has "Sam"
  "Jacob Rufe"            = 6011L,  # DB has "Jake"
  "Daniel Trejo"          = 6006L,  # DB has "Danny"
  "Preston Tabort Etaka"  = 6009L,  # name-part ordering differs in DB
  "RamizHamouda"          = 6004L,  # missing space in source
  "Axel Essenque"         = 6617L,  # typo variant of Axel Essengue
  "Remi Cabral"           = 6702L,  # accent-stripped variant of Rémi Cabral
  "Maximiliano Rodriguez" = 6685L,  # accent-stripped variant
  "Morey Doner"           = 6425L   # duplicate row in source; 6425 has data
)

# ── Resolve a CSV player name to a player_id ─────────────────────────────────
# Returns NA_integer_ if no unique match is found.
lookup_player_id <- function(csv_name, lookup_df) {
  # 1. Manual override
  if (csv_name %in% names(manual_overrides)) {
    id <- manual_overrides[[csv_name]]
    if (!is.na(id)) return(id)
  }
  # 2. Exact accent-normalised match
  norm <- stri_trans_general(csv_name, "Latin-ASCII")
  m    <- lookup_df[lookup_df$name_norm == norm, ]
  if (nrow(m) == 1) return(m$player_id)
  # 3. Last-name-only fallback (only if unique)
  last <- strsplit(norm, " ")[[1]]
  last <- last[length(last)]
  m2   <- lookup_df[grepl(last, lookup_df$name_norm, fixed = TRUE), ]
  if (nrow(m2) == 1) return(m2$player_id)
  NA_integer_
}

# ── Parse "M-D-YY" or "M-D-YYYY" date from a numbered CSV filename ───────────
parse_snap_date <- function(fname) {
  stem  <- sub("\\.csv$", "", basename(fname))
  parts <- strsplit(stem, "_")[[1]]
  bits  <- strsplit(parts[2], "-")[[1]]
  yr    <- as.integer(bits[3])
  if (yr < 100L) yr <- yr + 2000L
  as.Date(sprintf("%04d-%02d-%02d", yr, as.integer(bits[1]), as.integer(bits[2])))
}

# ── Map Excel row-label → player_snaps column name ───────────────────────────
stat_col_map <- c(
  "Games played"                            = "games_played",
  "Minutes played"                          = "mins_played",
  "Starts"                                  = "starts",
  "Subbed on"                               = "subbed_on",
  "Subbed off"                              = "subbed_off",
  "Goals conceded"                          = "goals_conceded",
  "Saves"                                   = "saves",
  "Catches"                                 = "catches",
  "Punches"                                 = "punches",
  "Goals"                                   = "goals",
  "Penalties"                               = "penalties",
  "Minutes per goal"                        = "mins_per_goal",
  "Shots on target"                         = "shots_on_target",
  "Shooting accuracy"                       = "shooting_accuracy",
  "Successful crosses"                      = "successful_crosses",
  "Crossing accuracy"                       = "crossing_accuracy",
  "Assists"                                 = "assists",
  "Key Passes"                              = "key_passes",
  "Penalties won"                           = "penalties_won",
  "Offsides"                                = "offsides",
  "Tackles won"                             = "tackles_won",
  "Tackles success rate"                    = "tackles_success_rate",
  "Clearances"                              = "clearances",
  "Blocks"                                  = "blocks",
  "Interceptions"                           = "interceptions",
  "Successful passes"                       = "successful_passes",
  "Passing accuracy"                        = "passing_accuracy",
  "Passing accuracy in opponents' half (%)" = "pass_accuracy_opp_half",
  "Successful dribbles"                     = "successful_dribbles",
  "Fouls won"                               = "fouls_won",
  "Fouls conceded"                          = "fouls_conceded",
  "Penalties conceded"                      = "penalties_conceded",
  "Yellow cards"                            = "yellow_cards",
  "Red cards"                               = "red_cards"
)
```

---

# 1 — Convert .xlsx to .csv

Skip this section if your files are already in `.csv` format.

```{r convert-xlsx}
xlsx_files <- list.files(base, pattern = "\\.xlsx$", full.names = TRUE,
                         recursive = TRUE)
cat("Found", length(xlsx_files), ".xlsx files\n")

converted <- 0L
for (f in xlsx_files) {
  csv_path <- sub("\\.xlsx$", ".csv", f)
  if (!file.exists(csv_path)) {
    df <- read_excel(f)
    write.csv(df, csv_path, row.names = FALSE)
    converted <- converted + 1L
  }
}
cat("Converted", converted, "new files (skipped", length(xlsx_files) - converted,
    "already-existing CSVs)\n")
```

---

# 2 — Ingest Named CSVs

These CSVs map directly to base tables: `matches`, `stadiums`, `teams`,
`players`, `appearances`, `player_snaps` (pre-season export), and
`season_team_stats`. New rows are inserted; existing rows are left unchanged.

```{r ingest-named-matches}
log_msg("--- matches ---")
csv <- read_csv(paste0(base, "matches.csv"), show_col_types = FALSE)
db  <- dbGetQuery(con, "SELECT match_id FROM matches")
new <- csv[!csv$match_id %in% db$match_id, ]
n   <- safe_insert(con, "matches", new)
log_msg("matches: inserted", n, "new rows (DB had", nrow(db), ")")
```

```{r ingest-named-stadiums}
log_msg("--- stadiums ---")
csv <- read_csv(paste0(base, "stadiums.csv"), show_col_types = FALSE)
db  <- dbGetQuery(con, "SELECT stadium_id FROM stadiums")
new <- csv[!csv$stadium_id %in% db$stadium_id, ]
n   <- safe_insert(con, "stadiums", new)
log_msg("stadiums: inserted", n, "new rows (DB had", nrow(db), ")")
```

```{r ingest-named-teams}
log_msg("--- teams ---")
csv <- read_csv(paste0(base, "teams.csv"), show_col_types = FALSE)
db  <- dbGetQuery(con, "SELECT team_id FROM teams")
# 'stadium' in the CSV corresponds to stadium_id; extra columns are dropped
new <- csv %>%
  rename(stadium_id = stadium) %>%
  select(team_id, name, city, abbreviation, conference, member_since, stadium_id) %>%
  filter(!team_id %in% db$team_id)
n <- safe_insert(con, "teams", new)
log_msg("teams: inserted", n, "new rows (DB had", nrow(db), ")")
```

```{r ingest-named-players}
log_msg("--- players ---")
csv <- read_csv(paste0(base, "players.csv"), show_col_types = FALSE,
                col_types = cols(dob = col_character()))
db  <- dbGetQuery(con, "SELECT player_id FROM players")
new <- csv %>%
  filter(!is.na(name)) %>%                   # drop all-NA placeholder rows
  filter(!player_id %in% db$player_id) %>%
  select(player_id, name, dob, nation, position, dom_foot)
n <- safe_insert(con, "players", new)
log_msg("players: inserted", n, "new rows (DB had", nrow(db), ")")
```

```{r ingest-named-appearances}
log_msg("--- appearances ---")
csv <- read_csv(paste0(base, "appearances.csv"), show_col_types = FALSE)
db  <- dbGetQuery(con, "SELECT match_id, player_id, team_id FROM appearances")
new <- csv %>%
  filter(!is.na(match_id), !is.na(player_id), !is.na(team_id)) %>%
  anti_join(db, by = c("match_id", "player_id", "team_id"))
n <- safe_insert(con, "appearances", new)
log_msg("appearances: inserted", n, "new rows (DB had", nrow(db), ")")
```

```{r ingest-named-player-snaps}
log_msg("--- player_snaps (pre-season CSV export) ---")
csv <- read_csv(paste0(base, "player_snaps.csv"), show_col_types = FALSE)
db  <- dbGetQuery(con, "SELECT snap_date, team_id, player_id FROM player_snaps")
db$snap_date <- as.Date(db$snap_date)
new <- csv %>%
  rename(pass_accuracy_opp_half = passing_accuracy_in_opponents_half) %>%
  filter(!is.na(player_id), !is.na(team_id)) %>%
  anti_join(db, by = c("snap_date", "team_id", "player_id")) %>%
  mutate(snapshot_id = NA_integer_)
n <- safe_insert(con, "player_snaps", new)
log_msg("player_snaps: inserted", n, "new rows (DB had", nrow(db), ")")
```

```{r ingest-named-season-team-stats}
log_msg("--- season_team_stats ---")
csv <- read_csv(paste0(base, "season_team_stats.csv"), show_col_types = FALSE)
if (nrow(csv) > 0) {
  new <- csv %>% rename(season = year)
  n   <- safe_insert(con, "season_team_stats", new)
  log_msg("season_team_stats: inserted", n, "rows")
} else {
  log_msg("season_team_stats: CSV is empty — nothing to import")
}
```

---

# 3 — Review New Player Names

The numbered CSVs (`100_8-23-25.csv`, etc.) are wide-format per-team
stat snapshots. Each column header is `"Player Name\nTeam Name"`. Before
importing, we need every player name to resolve to a `player_id`.

This section finds names that **cannot currently be matched** to any player
in the DB. Review the table, then edit the exclusion list in the next chunk.

## 3.1 Discover unmatched names

```{r find-unmatched}
numbered_files <- list.files(base, pattern = "^[0-9]+_[0-9].*\\.csv$",
                             full.names = TRUE)
cat("Numbered CSV files found:", length(numbered_files), "\n\n")

name_lookup <- build_name_lookup(con)
player_team_map <- list()   # csv_name → character vector of team labels

for (fpath in numbered_files) {
  df <- tryCatch(
    read_csv(fpath, show_col_types = FALSE,
             col_types = cols(.default = col_character())),
    error = function(e) NULL
  )
  if (is.null(df) || ncol(df) < 2) next

  for (pcol in names(df)[-1]) {
    parts    <- strsplit(pcol, "\n")[[1]]
    raw_name <- parts[1]
    team_lbl <- if (length(parts) >= 2) parts[2] else NA_character_

    # Skip Excel artifact names: "...N", numeric-only, or names ending in ...N
    if (grepl("\\.\\.\\.\\d+", raw_name) || grepl("^[0-9.]+$", raw_name)) next

    pid <- lookup_player_id(raw_name, name_lookup)
    if (is.na(pid)) {
      if (is.null(player_team_map[[raw_name]])) player_team_map[[raw_name]] <- character(0)
      if (!is.na(team_lbl)) {
        player_team_map[[raw_name]] <- unique(c(player_team_map[[raw_name]], team_lbl))
      }
    }
  }
}

unmatched_df <- data.frame(
  csv_name = names(player_team_map),
  teams    = sapply(player_team_map, paste, collapse = "; "),
  stringsAsFactors = FALSE
) %>% arrange(csv_name)

cat("Unmatched player names:", nrow(unmatched_df), "\n")
```

## 3.2 Inspect the unmatched names

Review the table below. Each row is a player name from a snapshot CSV that
does not match any existing `players` record. The `teams` column shows which
team(s) the player appeared under across all files.

```{r show-unmatched}
if (nrow(unmatched_df) == 0) {
  cat("All player names already matched. Nothing to add.\n")
} else {
  unmatched_df %>%
    kable(
      col.names = c("CSV Name", "Appeared Under Team(s)"),
      caption   = paste("Unmatched player names —", nrow(unmatched_df), "total")
    ) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    ) %>%
    scroll_box(height = "400px")
}
```

## 3.3 Exclude invalid names

**Edit the vector below** before running it. Add any names from the table
above that you do **not** want inserted into the `players` table. Common
reasons to exclude:

- The name looks like a data error (typo, test row, etc.)
- It is a known alias that should be added to `manual_overrides` instead of
  creating a new player record (e.g. "Sam" vs "Samuel" for an existing player)

```{r set-exclusions}
# ── EDIT THIS VECTOR ──────────────────────────────────────────────────────────
# Add names exactly as they appear in the "CSV Name" column above.
# Leave as character(0) to accept all unmatched names as new players.
names_to_exclude <- character(0)

# Example:
# names_to_exclude <- c(
#   "Test Player",          # obvious data error
#   "J Smith"              # ambiguous — add manual_override instead
# )
# ─────────────────────────────────────────────────────────────────────────────

approved_df <- unmatched_df %>%
  filter(!csv_name %in% names_to_exclude)

cat("Names approved for insertion:", nrow(approved_df), "\n")
cat("Names excluded:              ", nrow(unmatched_df) - nrow(approved_df), "\n")

if (length(names_to_exclude) > 0) {
  cat("\nExcluded:\n")
  for (nm in names_to_exclude) cat(" -", nm, "\n")
}
```

---

# 4 — Add Approved Players to the DB

Converts each approved "First Last" name to "Last, First" format and inserts
it into `players` with a new sequential `player_id`. All other fields (`dob`,
`nation`, `position`, `dom_foot`) are left `NULL` — fill them in manually or
via a future enrichment step.

```{r add-new-players}
if (nrow(approved_df) == 0) {
  cat("No new players to add.\n")
} else {
  max_id  <- dbGetQuery(con, "SELECT MAX(player_id) AS m FROM players")$m
  next_id <- max_id + 1L

  new_players <- data.frame(
    player_id = seq(next_id, next_id + nrow(approved_df) - 1L),
    name      = sapply(approved_df$csv_name, to_last_first),
    dob       = NA_character_,
    nation    = NA_character_,
    position  = NA_character_,
    dom_foot  = NA_character_,
    stringsAsFactors = FALSE
  )
  rownames(new_players) <- NULL

  dbWriteTable(con, "players", new_players, append = TRUE)
  log_msg("Inserted", nrow(new_players), "new players (IDs",
          min(new_players$player_id), "to", max(new_players$player_id), ")")

  new_players %>%
    select(player_id, name) %>%
    kable(col.names = c("New player_id", "Name (Last, First)"),
          caption = "Players added to DB") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
    scroll_box(height = "300px")
}
```

> **After inserting**, rebuild the name lookup so the ingestion step below
> picks up the new IDs.

```{r rebuild-lookup}
name_lookup <- build_name_lookup(con)
cat("Name lookup rebuilt —", nrow(name_lookup), "players total\n")
```

---

# 5 — Ingest Numbered Snapshot CSVs

Registers each file in `raw_snapshots` (one row per file) and transforms the
wide-format stat data into `player_snaps` rows (one row per player per date).

```{r ingest-snapshots, results='hold'}
existing_sources <- dbGetQuery(con, "SELECT source_file FROM raw_snapshots")$source_file

snap_inserted   <- 0L
ps_inserted     <- 0L
ps_skipped_dup  <- 0L
still_unmatched <- character(0)
insert_errors   <- list()

for (fpath in numbered_files) {
  fname   <- basename(fpath)
  stem    <- sub("\\.csv$", "", fname)
  parts   <- strsplit(stem, "_")[[1]]
  team_id <- as.integer(parts[1])
  snap_date <- tryCatch(parse_snap_date(fname), error = function(e) NA)

  if (is.na(snap_date)) {
    log_msg("SKIP (bad date):", fname)
    next
  }

  # ── Register in raw_snapshots ──────────────────────────────────────────────
  if (!fname %in% existing_sources) {
    dbExecute(con,
      "INSERT INTO raw_snapshots (team_id, snap_date, source_file, ingested_at)
       VALUES (?, ?, ?, ?)",
      list(team_id, as.character(snap_date), fname,
           format(Sys.time(), "%Y-%m-%d %H:%M:%S"))
    )
    snap_inserted <- snap_inserted + 1L
  }
  snapshot_id <- dbGetQuery(
    con,
    "SELECT snapshot_id FROM raw_snapshots WHERE source_file = ?",
    list(fname)
  )$snapshot_id

  # ── Parse wide → long ──────────────────────────────────────────────────────
  df <- tryCatch(
    read_csv(fpath, show_col_types = FALSE,
             col_types = cols(.default = col_character())),
    error = function(e) {
      log_msg("ERROR reading", fname, ":", e$message)
      NULL
    }
  )
  if (is.null(df) || ncol(df) < 2) next

  stat_names  <- df[[1]]
  player_cols <- names(df)[-1]

  for (pcol in player_cols) {
    raw_name <- strsplit(pcol, "\n")[[1]][1]

    # Skip artifact column names
    if (grepl("\\.\\.\\.\\d+", raw_name) || grepl("^[0-9.]+$", raw_name)) next

    player_id <- lookup_player_id(raw_name, name_lookup)
    if (is.na(player_id)) {
      still_unmatched <- unique(c(still_unmatched, raw_name))
      next
    }

    # Check for existing row (avoid duplicates)
    already <- dbGetQuery(
      con,
      "SELECT 1 FROM player_snaps WHERE snap_date=? AND team_id=? AND player_id=?",
      list(as.character(snap_date), team_id, player_id)
    )
    if (nrow(already) > 0) {
      ps_skipped_dup <- ps_skipped_dup + 1L
      next
    }

    # Map stat row-labels → column names; deduplicate duplicate stat rows
    vals        <- as.list(df[[pcol]])
    names(vals) <- stat_col_map[stat_names]
    vals        <- vals[!is.na(names(vals))]
    vals        <- vals[!duplicated(names(vals))]

    row_df <- as.data.frame(vals, stringsAsFactors = FALSE)
    num_cols <- setdiff(names(row_df), c("snap_date", "team_id", "player_id", "snapshot_id"))
    row_df[num_cols] <- lapply(row_df[num_cols],
                               function(x) suppressWarnings(as.numeric(x)))
    row_df$snap_date   <- as.character(snap_date)
    row_df$team_id     <- team_id
    row_df$player_id   <- player_id
    row_df$snapshot_id <- snapshot_id

    tryCatch(
      { dbWriteTable(con, "player_snaps", row_df, append = TRUE)
        ps_inserted <- ps_inserted + 1L },
      error = function(e) {
        insert_errors[[length(insert_errors) + 1]] <<-
          paste0(raw_name, " in ", fname, ": ", e$message)
      }
    )
  }
}

log_msg("raw_snapshots: inserted", snap_inserted, "new rows")
log_msg("player_snaps:  inserted", ps_inserted, "new rows")
log_msg("player_snaps:  skipped", ps_skipped_dup, "already-existing rows")
```

---

# 6 — Audit

## 6.1 Final row counts

```{r audit-counts}
tables_to_check <- c(
  "matches", "stadiums", "teams", "players", "appearances",
  "player_snaps", "season_team_stats", "raw_snapshots",
  "season_rosters", "season_staff"
)

count_df <- data.frame(
  table = tables_to_check,
  rows  = sapply(tables_to_check, function(t) {
    dbGetQuery(con, paste0("SELECT COUNT(*) AS n FROM ", t))$n
  }),
  stringsAsFactors = FALSE
)

count_df %>%
  kable(col.names = c("Table", "Row Count"),
        format.args = list(big.mark = ","),
        caption = "DB row counts after ingestion") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 6.2 Still-unmatched player names

Names that appeared in snapshot CSVs but could not be resolved to a
`player_id` even after this run. If this list is non-empty, investigate
whether each name needs a new player record or a `manual_overrides` entry.

```{r audit-unmatched}
if (length(still_unmatched) == 0) {
  cat("All player names successfully matched.\n")
} else {
  cat(length(still_unmatched), "names still unresolved:\n\n")
  data.frame(name = sort(still_unmatched)) %>%
    kable(col.names = "Unresolved CSV Name",
          caption = "Names with no matching player_id") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
}
```

## 6.3 Insert errors

Any rows that passed name-lookup but failed at the DB insert step.

```{r audit-errors}
if (length(insert_errors) == 0) {
  cat("No insert errors.\n")
} else {
  cat(length(insert_errors), "insert error(s):\n\n")
  for (e in insert_errors) cat(" !", e, "\n")
}
```

## 6.4 Snapshot coverage

How many distinct (team, date) snapshots are now in the DB, and which files
are registered in `raw_snapshots`?

```{r audit-coverage}
dbGetQuery(con, "
  SELECT rs.team_id, t.abbreviation AS team, rs.snap_date, rs.source_file
  FROM raw_snapshots rs
  LEFT JOIN teams t ON rs.team_id = t.team_id
  ORDER BY rs.team_id, rs.snap_date
") %>%
  kable(caption = "Registered snapshot files") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

---

# 7 — Cleanup

```{r disconnect}
dbDisconnect(con)
cat("Connection closed.\n")
```

---

# Appendix: Adding a manual override

If a CSV uses a name variant for a player **already in the DB** (nickname,
spelling error, missing accent, etc.) rather than adding a duplicate player
record, add an entry to `manual_overrides` in the **Setup** chunk:

```r
manual_overrides <- c(
  ...,
  "CSV Spelling Here" = 1234L   # the existing player_id
)
```

Then re-run from the **Setup** chunk onwards. The variant will be silently
resolved at lookup time and no new player row will be created.

---

# Appendix: Updating `stat_col_map`

If future snapshot exports add new stat rows, map them in `stat_col_map`
inside the **Setup** chunk:

```r
stat_col_map <- c(
  ...,
  "New stat label in Excel" = "new_column_in_player_snaps"
)
```

You will also need to `ALTER TABLE player_snaps ADD COLUMN new_column_in_player_snaps REAL;`
in the database before re-running ingestion.
