---
title: "FBref → SQLite Pipeline"
subtitle: "Getting started with worldfootballR and importing per-match player data"
author: "USL Championship Data Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5
)
```

---

# Overview

This document is a guided workflow for pulling USL Championship data from
[FBref.com](https://fbref.com) using the
[worldfootballR](https://jaseziv.github.io/worldfootballR/) R package, and
writing it into the local SQLite database.

**End goal:** populate the `matches` and `appearances` tables with real FBref
data so that any player's per-match statistics — minutes, goals, assists,
shots, passing, defensive actions, cards — are queryable in one place.

**Why worldfootballR instead of direct scraping?**
FBref is protected by Cloudflare and serves several data tables inside HTML
comments. worldfootballR handles both of those obstacles internally, and
adds built-in rate-limiting so your requests don't get blocked.

---

# 1. Installation and Setup

Run this block once per machine to install all required packages.

```{r install, eval=FALSE}
install.packages(c(
  "worldfootballR",   # FBref scraping
  "DBI",              # database interface
  "RSQLite",          # SQLite driver
  "tidyverse",        # data wrangling
  "stringdist"        # fuzzy name matching for player linking
))
```

```{r libraries}
library(worldfootballR)
library(DBI)
library(RSQLite)
library(tidyverse)
library(stringdist)
```

```{r connect}
# This file lives in ETL_02/; the database is one level up.
# Adjust the path if you move either file.
DB_PATH <- "../usl_championship_26.db"
con <- dbConnect(RSQLite::SQLite(), dbname = DB_PATH)
```

---

# 2. What Data Is Available for USL Championship

FBref covers USL Championship, but the depth of data depends on the season
and whether match officials submitted detailed reports.  Here is what to
realistically expect:

| Data type | FBref availability | Maps to DB table |
|---|---|---|
| Match results (score, date, teams) | ✅ Full coverage | `matches` |
| Squad season totals | ✅ Full coverage | `season_team_stats` |
| Player match summary (mins, goals, assists, shots, cards) | ✅ Most matches | `appearances` |
| Player match passing stats | ⚠️ Varies by match | `appearances` |
| Player match defensive actions | ⚠️ Varies by match | `appearances` |
| Player match advanced possession | ⚠️ Varies by match | — |
| Lineups / starters vs. substitutes | ✅ Most matches | `appearances.is_starter` |
| xG, progressive passes, carries | ⚠️ Limited for USL | — |

> **Note:** When a match has incomplete advanced stats, `fb_advanced_match_stats()`
> returns an empty data frame for that URL.  The pipeline below handles this
> gracefully by skipping empty results.

---

# 3. Orientation: Season Team Stats

This is the simplest call and a good sanity-check that your connection to
FBref is working before you run the heavier match-level pulls.

```{r team-stats, eval=FALSE}
# Pull standard squad stats for the current USL Championship season.
# stat_type options: "standard", "shooting", "passing", "defense",
#                    "possession", "playing_time", "misc", "keeper"
squad_stats <- fb_season_team_stats(
  country         = "USA",
  gender          = "M",
  season_end_year = 2026,
  tier            = "3rd",          # USL Championship = 3rd tier
  stat_type       = "standard"
)

glimpse(squad_stats)
```

```{r team-stats-demo, echo=FALSE}
# Placeholder output shown when eval=FALSE above.
# Replace with the live result after running the chunk above.
message("(Run the chunk above with internet access to see live FBref data.)")
```

Key columns returned: `Squad`, `Poss`, `Gls`, `Ast`, `xG`, `MP`, and the
FBref URL for each squad — which you will use in the next section.

---

# 4. Match Results and Match URLs

`fb_match_results()` returns every match for a season, including the
individual match page URLs.  Those URLs are the input for all per-match
player stat calls.

```{r match-results, eval=FALSE}
match_results <- fb_match_results(
  country         = "USA",
  gender          = "M",
  season_end_year = 2026,
  tier            = "3rd"
)

# Key columns: Date, Home, Away, HomeGoals, AwayGoals, MatchURL
glimpse(match_results)
```

```{r match-url-filter, eval=FALSE}
# Keep only completed matches (no MatchURL means the match hasn't
# been played yet or the report hasn't been filed).
completed <- match_results |>
  filter(!is.na(MatchURL), !is.na(HomeGoals))

cat("Completed matches with report URLs:", nrow(completed), "\n")

# Save URLs as a vector for the next steps.
match_urls <- completed$MatchURL
```

---

# 5. Per-Match Player Statistics

This is the core of the pipeline.  `fb_advanced_match_stats()` fetches a
single stat type for a batch of match URLs and returns one row per player
per match.

**Important:** FBref rate-limits requests.  Do not remove the `Sys.sleep()`
calls.  worldfootballR adds its own delay, but an explicit pause between
batches keeps you well within their tolerance.

## 5a. Summary stats (minutes, goals, assists, shots, cards)

```{r match-summary-stats, eval=FALSE}
# Pull in batches of 10 to be polite to FBref's servers.
batch_size <- 10
summary_list <- list()

for (i in seq(1, length(match_urls), by = batch_size)) {
  batch <- match_urls[i:min(i + batch_size - 1, length(match_urls))]

  result <- tryCatch(
    fb_advanced_match_stats(
      match_url       = batch,
      stat_type       = "summary",
      team_or_player  = "player"
    ),
    error = function(e) {
      message("Batch ", i, " error: ", e$message)
      NULL
    }
  )

  if (!is.null(result) && nrow(result) > 0) {
    summary_list[[length(summary_list) + 1]] <- result
  }

  Sys.sleep(3)   # pause between batches
}

player_summary <- bind_rows(summary_list)
glimpse(player_summary)
```

Key columns from `stat_type = "summary"`:
`Player`, `Team`, `Match_Date`, `MatchURL`, `Min`, `Gls`, `Ast`, `Sh`,
`SoT`, `CrdY`, `CrdR`, `Starting` ("Pitch" = starter, "Bench" = sub).

## 5b. Passing stats

```{r match-passing-stats, eval=FALSE}
passing_list <- list()

for (i in seq(1, length(match_urls), by = batch_size)) {
  batch <- match_urls[i:min(i + batch_size - 1, length(match_urls))]

  result <- tryCatch(
    fb_advanced_match_stats(
      match_url      = batch,
      stat_type      = "passing",
      team_or_player = "player"
    ),
    error = function(e) NULL
  )

  if (!is.null(result) && nrow(result) > 0) {
    passing_list[[length(passing_list) + 1]] <- result
  }

  Sys.sleep(3)
}

player_passing <- bind_rows(passing_list)
```

Key columns: `Cmp` (passes completed), `Att` (passes attempted),
`Cmp_pct`, `KP` (key passes), `PPA` (passes into penalty area).

## 5c. Defensive actions

```{r match-defense-stats, eval=FALSE}
defense_list <- list()

for (i in seq(1, length(match_urls), by = batch_size)) {
  batch <- match_urls[i:min(i + batch_size - 1, length(match_urls))]

  result <- tryCatch(
    fb_advanced_match_stats(
      match_url      = batch,
      stat_type      = "defense",
      team_or_player = "player"
    ),
    error = function(e) NULL
  )

  if (!is.null(result) && nrow(result) > 0) {
    defense_list[[length(defense_list) + 1]] <- result
  }

  Sys.sleep(3)
}

player_defense <- bind_rows(defense_list)
```

Key columns: `Tkl` (tackles), `TklW` (tackles won), `Int` (interceptions),
`Blocks`, `Clr` (clearances).

---

# 6. Joining and Cleaning the Fetched Data

Merge all three stat frames on the shared keys: `Player`, `Team`,
`MatchURL`.  Some matches won't have passing or defensive breakdowns —
those will be `NA` after the join, which is fine.

```{r join-stats, eval=FALSE}
# Minimal column selection avoids duplicate-column collisions.
summary_slim <- player_summary |>
  select(Player, Team, MatchURL, Match_Date,
         Min, Gls, Ast, Sh, CrdY, CrdR, Starting)

passing_slim <- player_passing |>
  select(Player, Team, MatchURL, Cmp, Att, KP) |>
  rename(passes_completed = Cmp, passes_attempted = Att, key_passes = KP)

defense_slim <- player_defense |>
  select(Player, Team, MatchURL, TklW, Int, Blocks, Clr) |>
  rename(tackles_won = TklW, interceptions = Int, blocks = Blocks,
         clearances = Clr)

player_match_stats <- summary_slim |>
  left_join(passing_slim, by = c("Player", "Team", "MatchURL")) |>
  left_join(defense_slim, by = c("Player", "Team", "MatchURL")) |>
  mutate(
    is_starter     = as.integer(Starting == "Pitch"),
    minutes_played = suppressWarnings(as.integer(Min)),
    goals          = replace_na(Gls, 0L),
    assists        = replace_na(Ast, 0L),
    shots          = replace_na(Sh,  0L),
    yellow_cards   = replace_na(CrdY, 0L),
    red_cards      = replace_na(CrdR, 0L),
    passes         = replace_na(passes_completed, NA_integer_),
    interceptions  = replace_na(interceptions, NA_integer_)
  ) |>
  select(-Min, -Gls, -Ast, -Sh, -CrdY, -CrdR, -Starting)

glimpse(player_match_stats)
```

---

# 7. Linking FBref Data to Your Database

FBref uses its own player/team names which may not match your DB exactly.
This section resolves both.

## 7a. Match FBref team names → `team_abv`

```{r team-linking, eval=FALSE}
db_teams <- dbGetQuery(con, "SELECT team_abv, name FROM teams")

# Normalize: lowercase, strip punctuation
norm <- function(x) tolower(gsub("[^a-z0-9 ]", "", x))

db_teams <- db_teams |>
  mutate(name_norm = norm(name))

fbref_teams <- player_match_stats |>
  distinct(Team) |>
  mutate(team_norm = norm(Team))

# Fuzzy join: pick the closest DB team name for each FBref team name
team_map <- fbref_teams |>
  rowwise() |>
  mutate(
    best_match_idx = which.min(stringdist(team_norm, db_teams$name_norm,
                                          method = "jw")),
    team_abv       = db_teams$team_abv[best_match_idx],
    matched_name   = db_teams$name[best_match_idx],
    distance       = min(stringdist(team_norm, db_teams$name_norm,
                                    method = "jw"))
  ) |>
  ungroup()

# Review matches with distance > 0.15 — those may need manual correction.
team_map |> filter(distance > 0.15) |> print()
```

> **Review the flagged rows** before proceeding.  You can correct them by
> adding explicit overrides:
> ```r
> team_map <- team_map |>
>   mutate(team_abv = case_when(
>     Team == "Some FBref Name" ~ "CORRECT_ABV",
>     TRUE ~ team_abv
>   ))
> ```

## 7b. Link FBref player names → `player_id`

The same fuzzy-matching approach is used for players, scoped within each
team's roster so a common name like "David Smith" on different teams doesn't
collide.

```{r player-linking, eval=FALSE}
db_players <- dbGetQuery(con, "
  SELECT p.player_id, p.name,
         sr.team_abv, sr.season
  FROM players p
  JOIN season_rosters sr ON p.player_id = sr.player_id
  WHERE sr.season = 2026
") |>
  mutate(name_norm = norm(name))

# Add team_abv to the FBref stats using the team map built above.
player_match_stats <- player_match_stats |>
  left_join(select(team_map, Team, team_abv), by = "Team") |>
  mutate(player_norm = norm(Player))

# For each row in the FBref data, find the closest player on that team's roster.
link_player <- function(player_norm_val, team_abv_val, db_players) {
  candidates <- db_players |> filter(team_abv == team_abv_val)
  if (nrow(candidates) == 0) return(NA_integer_)
  distances <- stringdist(player_norm_val, candidates$name_norm, method = "jw")
  best <- which.min(distances)
  if (distances[best] > 0.20) return(NA_integer_)  # threshold: no confident match
  candidates$player_id[best]
}

player_match_stats <- player_match_stats |>
  rowwise() |>
  mutate(player_id = link_player(player_norm, team_abv, db_players)) |>
  ungroup()

# How many rows were matched?
cat("Matched:", sum(!is.na(player_match_stats$player_id)),
    "/", nrow(player_match_stats), "player-match rows\n")

# Unmatched rows — review and add to players table if legitimate new players.
unmatched <- player_match_stats |> filter(is.na(player_id))
print(unmatched |> select(Player, Team, team_abv) |> distinct())
```

---

# 8. Writing to the Database

## 8a. Insert new matches into `matches`

FBref match results must be written to `matches` first because
`appearances.match_id` is a foreign key referencing it.

```{r write-matches, eval=FALSE}
# Map the FBref match results frame (from Section 4) to the DB schema.
existing_match_urls <- dbGetQuery(con,
  "SELECT fbref_url FROM matches WHERE fbref_url IS NOT NULL"
)$fbref_url

new_matches <- completed |>
  filter(!MatchURL %in% existing_match_urls) |>
  left_join(select(team_map, Team, team_abv), by = c("Home" = "Team")) |>
  rename(home_team_abv = team_abv) |>
  left_join(select(team_map, Team, team_abv), by = c("Away" = "Team")) |>
  rename(away_team_abv = team_abv) |>
  transmute(
    date          = as.character(Date),
    home_team_abv = home_team_abv,
    away_team_abv = away_team_abv,
    home_score    = as.integer(HomeGoals),
    away_score    = as.integer(AwayGoals),
    competition   = "USL Championship",
    fbref_url     = MatchURL          # store URL so we can skip re-imports
  )

# NOTE: your current matches schema doesn't have a fbref_url column yet.
# Add it once with:
#   dbExecute(con, "ALTER TABLE matches ADD COLUMN fbref_url TEXT")
# Then this insert will work cleanly on subsequent runs.

dbAppendTable(con, "matches", new_matches)
cat("Inserted", nrow(new_matches), "new match rows.\n")
```

## 8b. Insert player appearances

```{r write-appearances, eval=FALSE}
# Fetch the match_ids just inserted (needed to link appearances).
db_matches <- dbGetQuery(con,
  "SELECT match_id, fbref_url FROM matches WHERE fbref_url IS NOT NULL"
)

# Check which (match_id, player_id) pairs are already in appearances
# so we don't double-insert on re-runs.
existing_app <- dbGetQuery(con,
  "SELECT match_id, player_id FROM appearances"
)

new_appearances <- player_match_stats |>
  filter(!is.na(player_id), !is.na(team_abv)) |>
  left_join(db_matches, by = c("MatchURL" = "fbref_url")) |>
  filter(!is.na(match_id)) |>
  anti_join(existing_app, by = c("match_id", "player_id")) |>
  transmute(
    match_id       = match_id,
    player_id      = player_id,
    team_abv       = team_abv,
    minutes_played = minutes_played,
    is_starter     = is_starter,
    goals          = goals,
    shots          = shots,
    passes         = passes,
    assists        = assists,
    interceptions  = interceptions,
    duels_won      = NA_integer_,    # not in FBref summary; extend if needed
    yellow_cards   = yellow_cards,
    red_cards      = red_cards
  )

dbAppendTable(con, "appearances", new_appearances)
cat("Inserted", nrow(new_appearances), "appearance rows.\n")
```

---

# 9. Querying the Result

Once both tables are populated, a player's per-match performance is a
straightforward join:

```{r query-player-matches, eval=FALSE}
player_name_search <- "Ropapa Mensah"   # change to any player

dbGetQuery(con, "
  SELECT
    m.date,
    home.name  AS home_team,
    away.name  AS away_team,
    m.home_score,
    m.away_score,
    a.minutes_played,
    a.is_starter,
    a.goals,
    a.assists,
    a.shots,
    a.passes,
    a.interceptions,
    a.yellow_cards,
    a.red_cards
  FROM appearances a
  JOIN matches  m    ON a.match_id       = m.match_id
  JOIN teams    home ON m.home_team_abv  = home.team_abv
  JOIN teams    away ON m.away_team_abv  = away.team_abv
  JOIN players  p    ON a.player_id      = p.player_id
  WHERE p.name = ?
  ORDER BY m.date
", params = list(player_name_search))
```

---

# 10. Projected Full Pipeline (Repeatable Workflow)

Once the initial load is complete, refreshing with new match data is a
five-step process.  Run this sequence after each matchday.

```{r full-pipeline, eval=FALSE}
# ── Step 1: Pull current season match results ──────────────────────────────
match_results <- fb_match_results(
  country = "USA", gender = "M", season_end_year = 2026, tier = "3rd"
)

completed <- match_results |> filter(!is.na(MatchURL), !is.na(HomeGoals))

# ── Step 2: Identify only new matches not yet in the DB ───────────────────
existing_urls <- dbGetQuery(con,
  "SELECT fbref_url FROM matches WHERE fbref_url IS NOT NULL"
)$fbref_url

new_urls <- setdiff(completed$MatchURL, existing_urls)
cat(length(new_urls), "new match(es) to process.\n")

if (length(new_urls) == 0) {
  message("Database is already current — nothing to import.")
  stop()
}

# ── Step 3: Fetch player stats for new matches only ───────────────────────
fetch_stat <- function(urls, stat_type, batch = 10, pause = 3) {
  out <- list()
  for (i in seq(1, length(urls), by = batch)) {
    b <- urls[i:min(i + batch - 1, length(urls))]
    res <- tryCatch(
      fb_advanced_match_stats(b, stat_type = stat_type,
                               team_or_player = "player"),
      error = function(e) NULL
    )
    if (!is.null(res) && nrow(res) > 0) out[[length(out) + 1]] <- res
    Sys.sleep(pause)
  }
  bind_rows(out)
}

raw_summary  <- fetch_stat(new_urls, "summary")
raw_passing  <- fetch_stat(new_urls, "passing")
raw_defense  <- fetch_stat(new_urls, "defense")

# ── Step 4: Clean, join, and link to DB ids  ──────────────────────────────
# (Reuse the joining and linking code from Sections 6 and 7.)
# team_map and db_players should be rebuilt here if running as a
# standalone script rather than continuing a session.

# ── Step 5: Write to DB ───────────────────────────────────────────────────
# (Reuse the dbAppendTable blocks from Section 8.)

dbDisconnect(con)
message("Pipeline complete.")
```

---

# 11. Schema Note: `fbref_url` Column

The pipeline above stores each FBref match URL in a `fbref_url` column on
the `matches` table.  This is what makes incremental updates safe — the
script skips any match already in the DB.  Add the column once:

```{r add-fbref-url-col, eval=FALSE}
dbExecute(con, "ALTER TABLE matches ADD COLUMN fbref_url TEXT")
```

This is non-destructive and does not affect any existing rows.

---

# 12. Disconnect

```{r disconnect}
if (exists("con") && dbIsValid(con)) {
  dbDisconnect(con)
  message("Database connection closed.")
}
```

---

*worldfootballR documentation: https://jaseziv.github.io/worldfootballR/*
*FBref USL Championship: https://fbref.com/en/comps/73/USL-Championship-Stats*
